---
title: "miaTime: Microbiome Time Series Analysis"
date: "`r Sys.Date()`"
package: 
    miaTime
output: 
    BiocStyle::html_document:
        fig_height: 7
        fig_width: 10
        toc: yes
        toc_depth: 2
        number_sections: true
vignette: >
  %\VignetteIndexEntry{miaTime}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false

library(knitr)
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    cache = TRUE
)
```

## Introduction

`miaTime` is a package in the `r BiocStyle::Biocpkg("mia")` family, providing
tools for time series manipulation using the
`r BiocStyle::Biocpkg("TreeSummarizedExperiment")` data container.

## Installation

`miaTime` is hosted on Bioconductor, and can be installed using via
`BiocManager`.

```{r}
#| label: install
#| eval: false

BiocManager::install("miaTime")
```

## Load the package

Once installed, `miaTime` is made available in the usual way.

```{r}
#| label: load_package

library(miaTime)
```

## Divergence between time points

`miaTime` offers functions to calculate divergences. These can be calculated
based on samples and their corresponding base time point, e.g., first sample of
time series. Moreover, divergences can be calculated in rolling basis meaning
that a sample is compared to previous ith sample.

Divergences can be calculated with `get*Divergence()` functions. In the example
below, for each subject, we calculate the divergence of their samples by
comparing them to the first time point.

```{r}
#| label: base_divergence

data(hitchip1006)
tse <- hitchip1006

res <- getBaselineDivergence(tse, group = "sample", name = "baseline")
res |> head()
```

A more convenient and preferred approach is to store the values directly in
`colData` using the `get*Divergence()` functions. In the example below, we
calculate stepwise divergences with a lag of 1, meaning that for each sample,
the divergence is calculated by comparing it to the previous time point for
the same subject.

```{r}
#| label: time_divergence

tse <- addStepwiseDivergence(tse)
colData(tse)
```

## Visualize time series

We can visualile time series data with `r BiocStyle::Biocpkg("miaViz")`. Below,
we visualize 2 most abundant taxa.

```{r}
#| label: plot_series

# Transform data in relative scale
tse <- transformAssay(tse, method = "relabundance")
p <- plotSeries(
    tse, assay.type = "relabundance", x = "time", y = getTop(tse, 5))
p
```

See [articles](https://microbiome.github.io/miaTime/articles/) for more detailed
example workflows.

## Session info

```{r}
#| label: session_info

sessionInfo()
```
