---
title: "Time series manipulation"
date: "`r Sys.Date()`"
package: 
    miaTime
output: 
    BiocStyle::html_document:
        fig_height: 7
        fig_width: 10
        toc: yes
        toc_depth: 2
        number_sections: true
vignette: >
    %\VignetteIndexEntry{miaTime}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r}
#| label: setup
#| include: false

library(knitr)
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    cache = TRUE,
    message = FALSE,
    warning = FALSE,
    echo = TRUE
)
```

## Introduction

`miaTime` implements tools for time series manipulation based on the
`r BiocStyle::Biocpkg("TreeSummarizedExperiment")` (`TreeSE`) data container.
Much of the functionality is also applicable to the
`r BiocStyle::Biocpkg("SummarizedExperiment")` data objects. This tutorial
shows how to use `miaTime` methods as well as the broader R/Bioconductor
ecosystem to manipulate time series data.

Check also the related package
[TimeSeriesExperiment](https://github.com/nlhuong/TimeSeriesExperiment).

## Installation

`miaTime` is hosted on Bioconductor, and can be installed using via
`BiocManager`.

```{r}
#| label: install
#| eval: false

BiocManager::install("miaTime")
```

Once installed, `miaTime` is made available in the usual way.

```{r}
#| label: load_package

library(miaTime)
```

## Sorting samples 

To sort data based on subject and time point in base R, you can use the
`order()` function.

```{r}
#| label: sort_samples

data(hitchip1006)
tse <- hitchip1006

index <- order(tse[["subject"]], tse[["time"]])
tse <- tse[ , index]
```

## Storing time information with `period` class

`miaTime` utilizes the functions available in the package `lubridate`
to convert time series field to `period` class object. This gives access to a 
number of readily available
[time series manipulation tools](https://cran.r-project.org/web/packages/lubridate/vignettes/lubridate.html).


Load example data:

```{r}
#| label: lubridate

# Load packages
library(lubridate)

# Time is given in days in the demo data.
# Convert days to seconds
time_in_seconds <- 60*60*24*tse[["time"]]
# Convert the time data to period class
seconds <- as.period(time_in_seconds, unit = "sec")
# Check the output
seconds |> tail()
```

## Conversion between time units

The time field in days is now shown in seconds. It can then be
converted to many different units using the `lubridate` package.

```{r}
#| label: hours

hours <- as.period(seconds, unit = "hour")
hours |> tail()
```

The updated time information can then be added to the
`SummarizedExperiment` data object as a new `colData` (sample data)
field.

```{r}
#| label: add_seconds

colData(tse)$time_sec <- seconds
colData(tse)
```

## Calculating time differences 

The `lubridate::as.duration()` function helps to specify time points as
duration.

```{r}
#| label: duration

duration <- as.duration(seconds)
duration |> tail()
```

The difference between subsequent time points can then be calculated.

```{r}
#| label: time_diff

time_diff <- diff(duration)
time_diff <- c(NA, time_diff)
time_diff |> tail()
```

The time difference from a selected point to the other time points
can be calculated as follows.

```{r}
#| label: time_diff2

# Difference from second time point
time_diff <- hours - sort(unique(hours))[[2]]
time_diff |> tail()
```

## Time point rank

Rank of the time points can be calculated by `rank` function provided in base R.

```{r}
#| label: rank

tse[["time"]] <- rank(tse[["time"]])
colData(tse)
```

## Operations per unit

Sometimes we need to operate on time series per unit (subject, reaction chamber,
sampling location, ...).

Add time point rank per subject.

```{r}
#| label: rank_for_subject

library(dplyr)

colData(tse) <- colData(tse) |>
   as.data.frame() |>
   group_by(subject) |>
   mutate(rank = rank(time, ties.method = "average")) |>
   DataFrame()
```

## Subset to baseline samples

`TreeSE` consists of rows for features and columns for samples. If we are
specifically interested in baseline samples, we can easily subset the data as
follows.

```{r}
#| label: subset

tse <- tse[, tse$time==0]
```

# Session info

```{r}
#| label: session_info

sessionInfo()
```
