---
title: "Minimal gut bioreactor examples"
date: "`r Sys.Date()`"
package: 
    miaTime
output: 
    BiocStyle::html_document:
        fig_height: 7
        fig_width: 10
        toc: yes
        toc_depth: 2
        number_sections: true
vignette: >
    %\VignetteIndexEntry{miaTime}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r}
#| label: setup
#| include: false

library(knitr)
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    cache = TRUE
)
```

# Minimal gut microbiome

Dense samples of the minimal gut microbiome. In the initial hours, MDb-MM was
grown under batch condition and 24 h onwards, continuous feeding of media with
pulse feeding cycles. This information is stored in the `colData`.   

```{r, message = FALSE, warning = FALSE}
#| label: sample_table
library(miaTime)

data(minimalgut)
tse <- minimalgut

# Quick check of number of samples 
table(tse[["StudyIdentifier"]], tse[["condition_1"]])
```

Visualize samples available for each of the bioreactors. This allows
to identify if there are any missing samples for specific times.

```{r, message = FALSE, warning = FALSE}
#| label: show_timepoints

library(ggplot2)

colData(tse) |>
    ggplot() +
    geom_tile(
        aes(x = as.factor(Time.hr), y = StudyIdentifier, fill = condition_1))
```

## Community dynamics   

The `minimalgut` dataset, mucus-diet based minimal microbiome
(MDbMM-16), consists of 16 species assembled in three bioreactors. We
can investigate the succession of mdbMM16 from the start of experiment
here hour zero until the end of the experiment.

```{r, message = FALSE, warning = FALSE}
#| label: calculate_divergence

# Transform data to relativeS
tse <- transformAssay(tse, method = "relabundance")
# Divergence from baseline i.e from hour zero
tse <- addBaselineDivergence(
    tse,
    assay.type = "relabundance",
    method = "bray",
    group = "StudyIdentifier",
    time.col = "Time.hr",
    )
```

Let's then visualize the divergence.

```{r, message = FALSE, warning = FALSE}
#| label: show_divergence

library(scater)

# Create a time series plot for divergence
p <- plotColData(
    tse, x = "Time.hr", y = "divergence", colour_by = "StudyIdentifier") +
    # Add line between points
    geom_line(aes(group = .data[["colour_by"]], colour =  .data[["colour_by"]]))
p
```

## Visualizing selected taxa

Now visualize abundance of _Blautia hydrogenotrophica_ using the
`miaViz::plotSeries()` function.

```{r, message = FALSE, warning = FALSE}
#| label: plot_series

library(miaViz)

# Plot certain feature by time
p <- plotSeries(
    tse,
    x = "Time.hr", y = "Blautia_hydrogenotrophica", colour_by = "Species",
    assay.type = "relabundance")
p
```

## Visualize the rate (slope) of divergence 

Sample dissimilarity between consecutive time steps(step size n >= 1) within
a group(subject, age, reaction chamber, etc.) can be calculated by
`addStepwiseDivergence`.

```{r, message = FALSE, warning = FALSE}
#| label: stepwise_divergence

# Divergence between consecutive time points
tse <- addStepwiseDivergence(
    tse,
    assay.type = "relabundance",
    method = "bray",
    group = "StudyIdentifier",
    time.interval = 1,
    time.col = "Time.hr",
    name = "divergence_from_previous_step",
    name.time = "time_from_previous_step"
    )
```

The results are again stored in `colData`. We calculate the speed of divergence
change by dividing each divergence change by the corresponding change in time.
Then we use similar plotting methods as previously.

```{r, message = FALSE, warning = FALSE}
#| label: show_stepwise

# Calculate slope for the change
tse[["divergence_change"]] <- tse[["divergence_from_previous_step"]] /
    tse[["time_from_previous_step"]]

# Create a time series plot for divergence
p <- plotColData(
    tse,
    x = "Time.hr",
    y = "divergence_change",
    colour_by = "StudyIdentifier"
    ) +
    # Add line between points
    geom_line(aes(group = .data[["colour_by"]], colour =  .data[["colour_by"]]))
p
```

## Moving average of the slope

This shows how to calculate and plot moving average for the variable of
interest (here: slope).

```{r, message = FALSE, warning = FALSE}
#| label: moving_average

library(dplyr)

# Calculate moving average with time window of 3 time points
tse[["sliding_divergence"]] <- colData(tse) |>
    as.data.frame() |>
    # Group based on reactor
    group_by(StudyIdentifier) |>
    # Calculate moving average
    mutate(sliding_avg = (
        # We get the previous 2 samples
        lag(divergence_change, 2) +
        lag(divergence_change, 1) +
        # And the current sample
        divergence_change
        # And take average
        ) / 3
    ) |>
    # Get only the values as vector
    ungroup() |>
    pull(sliding_avg)
```

After calculating the moving average of divergences, we can visualize the
result in a similar way to our previous approach.

```{r, message = FALSE, warning = FALSE}
#| label: show_moving_average

# Create a time series plot for divergence
p <- plotColData(
    tse,
    x = "Time.hr",
    y = "sliding_divergence",
    colour_by = "StudyIdentifier"
    ) +
    # Add line between points
    geom_line(aes(group = .data[["colour_by"]], colour =  .data[["colour_by"]]))
p
```

# Session info

```{r}
#| label: session_info

sessionInfo()
```
